Unsupervised Learning for Identifying Events in Active Target Experiments

### 2021-03-13 reviewer_responses.tex

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{Dear Editors,}

We are much indebted for the positive feedback from the reviewer and your willingness to consider our manuscript for publication.
We have tried as faithfully as possible to answer the referee and we hope the revised version meets your standards for publications in NIMA.\newline
Our answers to the referee are included below here. The revised version is included as well.\newline
Yours sincerely,\newline
Daniel Bazin, Morten Hjorth-Jensen, Michelle Kuchera, Robert Solli, and Ryan Strauss

@@

\begin{itemize}
\item Reviewer comment: Line 29, Simulated data? Line 193 simulated events… from a Ar(p,p) experiment. Are these
simulation data or experimental data? The use of word simulated is confusing. Based on the context,
it appears that only experimental data is used for this work.\\
\textcolor{blue}{Our answer: We apologize for the confusion. Both simulated events and records from the AT-TPC were studied. Simulated events were generated from an event simulator written for the purpose of analysing the Ar(p,p) reaction. We added a sentence in the abstract to clarify the circumstance. }
\item Reviewer comment: Line 124-125, simulated via an integration, which is numerically costly. Rephrase and clarify how
costly quantitatively. Similar clarifications needed in Line 131, Line 145 and Table 2.\\
\textcolor{blue}{Our answer: More quantitative information about the computational cost of traditional analysis methods was added in the paragraphs mentioned by the referee.}
\item Reviewer comment: Figure 1. Add some discussions about the experimental information and physics behind the
observed patterns. For example, which direction is the incoming beam? Beam energy? Vertex at
(0,0)? The direction of the magnetic field?\\

\textcolor{blue}{Our answer: Additional text has been added in the figure caption to better explain the experimental conditions of the displayed data.}
\item Reviewer comment: Line 306, Sample $\hat{x}$, presumably each of these is a 2D image x-y? what are the sample
dimensions?

\textcolor{blue}{Our answer: The text has been modified to clarify that the samples are images encoded as a square matrix with 128 elements along one axis.}

\item Reviewer comment: Lines 330 \& 331, define Ground truth labels y and predictions $\hat{y}$, how are they derived from
the raw data (2D projections of the AT-TPC)? Or the latent layer outputs of a pre-trained neural
network? Or else?
\textcolor{blue}{Our answer: We have added more text in each subsection of Section 3: Data Preparation was added to indicate where the ground truth labels were derived from. In fact, we used this data set {\it because} we had rare access to ground truth labels for experimental data.}
\item Reviewer comment: Figures 4, Figure 5, Figure 6, Figure 9, Figure 10, are data science metric results. Needs update or
move some of them to the appendix.
\textcolor{blue}{Our answer: We have followed the reviewer's suggestions and have moved the confusion matrix figures to the appendix, and we added text to clarify their meaning.}
\item Reviewer comment: Figure 7, Figure 8, elucidate the physics and understanding behind the pattern would be helpful.
Similar to the comments to Figure 1 above.
\textcolor{blue}{Our answer: These figures are now the new Figures 4 and 5. Since this is unsupervised clustering, it is normally difficult to conjecture a clear physical meaning behind the separation. However, the plots visualize a sampling of proton events that were attributed to a noisy and pure cluster (each cluster), allowing a reader to infer the behavior of the unsupervised learning on these events. The text has been updated with intuitive physics deductions from the plots. The figure caption texts have also been updated for clarity. }
\end{itemize}

### 2021-03-13 doc/Manuscript/article.tex
v2 - revision submission

%\documentclass[preprint,12pt]{elsarticle}
% uncomment for submission to NIM A
\documentclass[review,sort&compress]{elsarticle}
\usepackage{lineno}
\linenumbers

@@

\begin{abstract}
% Modified by DB (6/8/2020)
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labeled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
%With autoencoder neural networks we find improved descriptions of data from experiments.
This article presents novel applications of unsupervised machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to group similar events in the early stages of the data analysis, thereby improving efficiency by limiting the computationally expensive processing of unnecessary events.
The application of unsupervised clustering algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network. We study clustering performance on data from a simulated $^{46}$Ar experiment, and records from the AT-TPC detector. 
We find that a $K$-means algorithm applied to simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+$K$-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of clustering the latent space of autoencoder neural networks for event separation. While these networks show strong performance, they suffer from high variability in their results.

@@

A set of $N=8000$ simulated AT-TPC events per class was generated from the \texttt{pytpc} package developed by \citet{Bradt2017a} for the analysis of the  ${}^{46}$Ar$(p, p)$ experiment.
%A set of $N=8000$ simulated AT-TPC events per class was obtained  from a ${}^{46}$Ar$(p, p)$ experiment with the \texttt{pytpc} package developed by \citet{Bradt2017a}.%, see table \ref{tab:class_distr}. %Using the same parameters {\color{blue} what parameters are you referring to? -MPK} as for the 
%Two simulated ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.

For validation, we select a subset of the simulated data to be labeled and treat the rest as unlabeled data. We chose this partition to consist of  $15\%$ of the data. We denote this subset and its associated labels as $\gamma_L=(\boldsymbol{X}_L, \boldsymbol{y}_L)$, while the entire data set is identified as $\boldsymbol{X}_F$. Note that $\boldsymbol{X}_L \subset \boldsymbol{X}_F$.

{\bf As this dataset is generated via simulation, the true labels are known for this dataset.}

@@

The events analyzed in this section were retrieved from the ${}^{46}$Ar resonant proton scattering experiment recorded with the AT-TPC. 
While we denote these events as raw, it is important to note that what we intend is a raw two-dimensional projection of events {\bf without any post-processing done to clean the data}.

@@

We display two different events from the ${}^{46}$Ar experiment in Fig.~\ref{fig:samples}. The top row illustrates a proton event with a distinctive spiral-shaped track, while the bottom row shows a carbon event with a large fraction of noise.

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{custom_work/examples_raw.pdf}
\caption[Displaying unfiltered events in two and three dimensions]{Two- and three-dimensional representations of two events from the ${}^{46}$Ar experiment. Each row is one event in two projections, where the color intensity of each point indicates higher charge values recorded by the detector. The top row illustrates a proton event with a distinctive spiral-shaped track, while the bottom row shows a carbon event with a large fraction of noise. The beam direction is oriented along the z-axis (from positive to negative), close to the (x,y) origin. The magnetic field is oriented on the same axis as the beam, causing the bending of the charged particle's trajectories. The spiral shape is caused by the slowing down of the particles inside the gas volume. The bending radius of the tracks depends not only on the energy, but also on the mass-to-charge ratio of the particles, hence the large difference between proton and carbon events (see also section \ref{Sec:Exp}).}
\label{fig:samples}
\end{figure}

{\bf A subset of this data was hand-labeled for validation.}
\subsection{Filtered \texorpdfstring{${}^{46}$Ar}{46Ar} events}\label{sec:filtered}

@@

We illustrate two filtered events in Fig.~\ref{fig:samples_filtered}. These are the same events as shown in Fig.~\ref{fig:samples}, but with the Hough transform and nearest-neighbor filtering applied.

\begin{figure}[ht]
\centering
\includegraphics[width=0.9\textwidth, height=9cm]{custom_work/examples_filtered.pdf}
\caption[Displaying filtered events in 2D and 3D]{Two- and three-dimensional representations of two events from the ${}^{46}$Ar experiment. Each row is one event in two projections, where the lightness of each point indicates higher charge values. These events have been filtered with a nearest neighbors algorithm and an algorithm based on the Hough transform \cite{Newman1972}.}\label{fig:samples_filtered}
\end{figure}
{\bf The same subset of hand-labeled data from Section~\ref{sec:data_real} was used for validation.}

@@

\subsection{Classifying  events} 
%The filtering task relied on a set of $\chi^2$ distributions based on different criteria, 
The traditional Monte Carlo event selection process, described in Section~\ref{Sec:Exp}, does not have a well-defined method to quantify the effectiveness of the event selection.
In addition, the selection task produced a binary result only, either a {\em good} or {\em bad} fit to the event of interest. A {\em bad} fit is then assumed to be a different event type, and is removed from the analysis. %, if only because of the presence of carbon in the target (from the isobutane gas).
In a broader perspective, an unsupervised classification algorithm would offer the possibility to {\em discover} rare events which may not be expected or are overlooked. These events would likely be filtered out using the traditional methods. From a practical point of view, compared to supervised learning, it also avoids the necessary labeling task of the learning set events, which is error prone and time consuming.

@@

The $\chi^2$ approach used in the traditional analysis performed on the $^{46}$Ar data is extremely expensive from a computational stand because it involves the simulation of thousands of tracks for each recorded event. These events are in turn simulated for each iteration of the Monte Carlo fitting sequence. Even though the reaction of interest in the $^{46}$Ar experiment had the largest cross section (elastic scattering), the time spent on Monte Carlo fitting of {\em all} of the events produced in the experiment was the largest computational bottleneck in the analysis. In the case of an experiment where the reaction of interest would represent less than a few percent of the total cross section, this procedure would become highly inefficient and prohibitive. Adding to this the large amount of data produced in this experiment (with even larger data sets expected in future experiments), the analysis simply begs for more efficient analysis tools. 
%Most of the CPU time would be spent on simulating events that are not of interest.
The computationally expensive fitting procedure would be applied to every event, instead of the few percent of the events that are of interest for the analysis.
An unsupervised ML algorithm able to separate the data without {\em a priori} knowledge of the different types of events increases the efficiency of the analysis tremendously, and allows the downstream analysis to concentrate on the fitting efforts only on events of interest. In addition, the clustering allows for more exploration of the data, potentially enabling new discovery of unexpected reaction types.

@@

% The clustering algorithm we found to give the strongest result is the $K$-means algorithm. This algorithm is contingent on having a distance measurement between two datapoints. We found that the standard choice of a euclidian norm on the vector differences provided the best outcomes.

In contrast with the classification work of \citet{Kuchera2019}, we do not assume access to ground truth labels and we are trying to solve a fundamentally different learning problem. Thus, rather than fine-tuning a pre-trained network under the supervised learning regime, we extract the output of the pre-trained network's last convolutional layer as a latent representation of the events, where each event is represented as a vector in $\R^{8192}$. We then cluster events based on this representation using the Scikit-Learn implementation of the $K$-means algorithm with default parameters \cite{Pedregosa2011}.

%{\color{cyan} [Ryan: I think there are a lot of words in the following two paragraphs that don't need to exist. Essentially, these paragraphs are a repeat of the last paragraph and parts of the last subsection, but with one or two extra details added in (e.g. using the top 100 principle components). As such, I would remove them and add the necessary details to the previous paragraph (or replace them with a small paragraph).]} Building on the work by \citet{Kuchera2019}, we also use the VGG16 pre-trained network as a baseline for the clustering performance. We begin by considering a classical $K$-means approach to clustering. However, the output from the VGG16 network is very high dimensional which traditionally conflicts with good $K$-means results. With output vectors in $\R^{8192}$ the ratio between $L_p$ distances goes to one with increasing dimensionality, as shown by \citet{Aggarwal}. However, one of the central assumptions in the authors' finding is that the elements are uniformly distributed in the latent space. They note that it is then possible that if the class information lies in some sub-space of the latent data, $L_p$-norms may provide useful clusterings even in high dimensional spaces. To investigate this, we perform a $L_2$-norm based $K$-means clustering using the full VGG16 representation and the $10^2$ first principal components only.

%{\color{blue} Are there supposed to be results here? -MPK}
%\subsection{$K$-means}

%We begin by investigating the $K$-means clustering algorithm on the VGG16 latent space. 
% The user a VGG16 model that is pre-trained on the ImageNet data set and subsequently applied to the AT-TPC events without the classification layers. This creates a set of vectors $\boldsymbol{x} \in \R^{8192}$ as our new representation of the events. To cluster, we use the \lstinline{scikit-learn} implementation of the $K$-means++ algorithm with default parameters \cite{Pedregosa2011}.

@@

To compute both the ARI and the clustering accuracy we also have to construct the contingency table between two sets of clustering assignments. A contingency matrix defines the overlaps between classes in these sets, and its general form is shown in table \ref{tab:contingency}.

@@

\begin{equation}\label{eq:ari}
\text{ARI} = \frac{\sum  \binom{w_{ij}}{2} - \left[\sum  \binom{e_{i}}{2} \sum  \binom{f_{j}}{2}  \right]/\binom{n}{2}}{\frac{1}{2}\left[\sum  \binom{e_{i}}{2} + \sum  \binom{f_{j}}{2}  \right]- \left[\sum  \binom{e_{i}}{2} \sum  \binom{f_{j}}{2}  \right]/\binom{n}{2}}.
\end{equation}
The quantities $f_i$ and $e_i$ are defined in table \ref{tab:contingency}.
The important distinction to make between the clustering accuracy and the ARI is that the clustering accuracy is a simple comparison that does not account for chance assignments, while the ARI does. In effect, this means that the accuracy is a good heuristic for the performance, tempered by the ARI.

@@

The performance is measured by comparing model predictions on the labelled subset of data (see table \ref{tab:data sets} in  appendix B for more details).

@@

We first explore the results of applying a $K$-means approach on the latent space of a pre-trained network. Subsequently, we investigate the performance of the MIXAE algorithm as outlined in section \ref{sec:mixae}.

\subsection{$K$-means clustering on the VGG16 latent space}\label{sec:kmeans_results}

The results of the clustering runs are included in table \ref{tab:clstr_vgg1}. We ran the $K$-means algorithm $N=100$ times with $M=10$ initializations per run, of the cluster centroids to assess the variability in the performance. The K-means algorithm returns the best performing model of the $M$ initializations on the unsupervised objective. We report performance on the labelled subset of data, using the labels to identify the top-performing model (Top 1). Additionally, we report the mean and standard deviation of the algorithm on the $N$ trials, which indicate unsupervised performance.

@@

We observe that the clustering on simulated data attains the highest performance, and that there is a decline in performance as we add noise by moving to the filtered and raw data sets. The results are shown in table \ref{tab:clstr_vgg10}.

\begin{table}[H]
\centering 
\caption[$K$-means on pre-trained model]{$K$-means clustering results on AT-TPC event data in the VGG-16 latent space, for $N=100$ runs of the $K$-means algorithm with $M=10$ initializations. We observe that the performance predictably decreases with the amount of noise in the data.}\label{tab:clstr_vgg10}
\input{./plots/kmeans_vgg_init10.tex}
\end{table}

\noindent The lack of variability is explained by the number of initializations. As can be seen from table \ref{tab:clstr_vgg1} where we run the $K$-means algorithm $N=1000$ times with $M=1$ initializations of the centroids.

\begin{table}[H]
\centering 
\caption[$K$-means on pre-trained model]{$K$-means clustering results on AT-TPC event data in the VGG-16 latent space, for $N=1000$ runs of the $K$-means algorithm with $M=1$ initializations. We observe that there is significant variability in the results, which is ordinarily masked by $M$ re-initializations that avoid local minima. }\label{tab:clstr_vgg1}
\input{./plots/kmeans_vgg_init1.tex}
\end{table}

In addition to the performance measures reported in table \ref{tab:clstr_vgg1}, it is interesting to observe which samples are wrongly assigned. To investigate this problem, we compute the matrices $\mathbf{W}$ as shown in table \ref{tab:contingency}. From these tables, we can infer which classes are more or less entangled with others. The results for each data set is shown in  Figs.~\ref{fig:clster_confmat_sim}, \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. We observe that the proton class is consistently assigned in a pure cluster. For example, consider the row corresponding to the proton class in Fig.~\ref{fig:clster_confmati_filt}. The column corresponding to the largest entry in the proton row has zero other predicted classes in it.

@@

To further investigate the clusters presented in the matrix in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}, we study examples from the proton samples belonging to different clusters for the filtered and full data.
We display these examples in Figs.~\ref{fig:filtered_vgg_clster_repr} and \ref{fig:full_vgg_clster_repr}, respectively. Form these figures we observe that the more noisy proton events are being clustered together with the amorphous other class events. This is in line with the intuitive expectation from viewing the clustering results in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}.

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Filteredexamples.pdf}
\caption[Filtered proton samples by cluster belonging]{A sample of proton events from different $K$-means clusters from the filtered data set. The bottom row shows proton samples that are intermingled with noise events in the models' predictions, and the top row belongs to a cluster that contains proton events almost exclusively. Each row belongs to a single cluster corresponding to the filtered confusion matrix in Fig.~\ref{fig:clster_confmati_filt}}\label{fig:filtered_vgg_clster_repr}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Rawexamples.pdf}
\caption[Full proton samples by cluster belonging]{Illustration of  a sample of proton events from different $K$-means clusters originating from the raw data set. Each row belongs to a single cluster corresponding to the full confusion matrix in Fig.~\ref{fig:clster_confmati_raw}}\label{fig:full_vgg_clster_repr}
\end{figure}

@@

With the best set of hyperparameters, each highest performing model is thereafter run $10$ times\footnote{The MIXAE model is significantly more computationally expensive to train than a $K$-means model. Resulting in the skew in number of runs in the two cases.}. The results are listed in table \ref{tab:clstr_mixae}.
We observe that, while the algorithm can achieve a very strong performance, the performance varies. In some cases the MIXAE model converges to a seemingly good configuration, based on its unsupervised training goals. However, when inspecting its clustering performance against labelled data, the seemingly good model does no better than a model based on random selection. 
This happens more frequently with the raw data, indicating an interaction with the noise levels present in the events.

>>>>

\documentclass[preprint,12pt]{elsarticle}
% uncomment for submission to NIM A
%\documentclass[review,sort&compress]{elsarticle}
%\usepackage{lineno}
%\linenumbers

@@

\begin{abstract}
% Modified by DB (6/8/2020)
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labeled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
%With autoencoder neural networks we find improved descriptions of data from experiments.
This article presents novel applications of unsupervised machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to group similar events in the early stages of the data analysis, thereby improving efficiency by limiting the computationally expensive processing of unnecessary events.
The application of unsupervised clustering algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network. We study clustering performance on both data from a simulated $^{46}$Ar experiment, and real events from the AT-TPC detector. 
We find that a $k$-means algorithm applied to simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+$k$-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of clustering the latent space of autoencoder neural networks for event separation. While these networks show strong performance, they suffer from high variability in their results.

@@

A set of $N=8000$ simulated AT-TPC events per class was generated from the \texttt{pytpc} package developed by \citet{Bradt2017a} for the analysis of the  ${}^{46}$Ar$(p, p)$ experiment.
%A set of $N=8000$ simulated AT-TPC events per class was obtained  from a ${}^{46}$Ar$(p, p)$ experiment with the \texttt{pytpc} package developed by \citet{Bradt2017a}.%, see Table \ref{tab:class_distr}. %Using the same parameters {\color{blue} what parameters are you referring to? -MPK} as for the 
%Two simulated ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.

For validation, we select a subset of the simulated data to be labeled and treat the rest as unlabeled data. We chose this partition to consist of  $15\%$ of the data. We denote this subset and its associated labels as $\gamma_L=(\boldsymbol{X}_L, \boldsymbol{y}_L)$, while the entire data set is identified as $\boldsymbol{X}_F$. Note that $\boldsymbol{X}_L \subset \boldsymbol{X}_F$.
As this dataset is generated via simulation, the true labels are known for this dataset.

@@

The events analyzed in this section were retrieved from the ${}^{46}$Ar resonant proton scattering experiment recorded with the AT-TPC. 
While we denote these events as raw, it is important to note that what we intend is a raw two-dimensional projection of events without any post-processing done to clean the data.

@@

We display two different events from the ${}^{46}$Ar experiment in Fig.~\ref{fig:samples}. The top row illustrates a proton event with a distinctive spiral-shaped track, while the bottom row shows a carbon event with a large fraction of noise. 
A subset of this data was hand-labeled for validation, as discussed in \cite{Kuchera2019}.
\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{custom_work/examples_raw.pdf}
\caption[Displaying unfiltered events in two and three dimensions]{Two- and three-dimensional representations of two events from the ${}^{46}$Ar experiment. Each row is one event in two projections, where the color intensity of each point indicates higher charge values recorded by the detector. The top row illustrates a proton event with a distinctive spiral-shaped track, while the bottom row shows a carbon event with a large fraction of noise. The beam direction is oriented along the z-axis (from positive to negative), close to the (x,y) origin. The magnetic field is oriented on the same axis as the beam, causing the bending of the charged particle's trajectories. The spiral shape is caused by the slowing down of the particles inside the gas volume. The bending radius of the tracks depends not only on the energy, but also on the mass-to-charge ratio of the particles, hence the large difference between proton and carbon events (see also section \ref{Sec:Exp}).}
\label{fig:samples}
\end{figure}

\subsection{Filtered \texorpdfstring{${}^{46}$Ar}{46Ar} events}\label{sec:filtered}

@@

We illustrate two filtered events in Fig.~\ref{fig:samples_filtered}. These are the same events as shown in Fig.~\ref{fig:samples}, but with the Hough transform and nearest-neighbor filtering applied. 
The same subset of hand-labeled data from Section~\ref{sec:data_real} was used for validation.

\begin{figure}[ht]
\centering
\includegraphics[width=0.9\textwidth, height=9cm]{custom_work/examples_filtered.pdf}
\caption[Displaying filtered events in 2D and 3D]{Two- and three-dimensional representations of two events from the ${}^{46}$Ar experiment. Each row is one event in two projections, where the lightness of each point indicates higher charge values. These events have been filtered with a nearest neighbors algorithm and an algorithm based on the Hough transform \cite{Newman1972}.}\label{fig:samples_filtered}
\end{figure}

@@

\subsection{Classifying  events} 
%The filtering task relied on a set of $\chi^2$ distributions based on different criteria, 
The traditional Monte Carlo event selection process, described in Section~\ref{Sec:Exp}, does not have a well-defined method to quantify the effectiveness of the event selection.
In addition, the selection task produced a binary, cut-based result: either a {\em good} or {\em bad} fit to the event of interest. A {\em bad} fit is then assumed to be a different event type, and is removed from the analysis. %, if only because of the presence of carbon in the target (from the isobutane gas).
In a broader perspective, an unsupervised classification algorithm would offer the possibility to {\em discover} rare events which may not be expected or are overlooked. These events would likely be filtered out using the traditional methods. From a practical point of view, compared to supervised learning, it also avoids the necessary labeling task of the learning set events, which is error prone and time consuming.

@@

The $\chi^2$ approach used in the traditional analysis performed on the $^{46}$Ar data is extremely expensive computationally because it requires the simulation of thousands of tracks for each recorded event. These events are in turn simulated for each iteration of the Monte Carlo fitting sequence. Even though the reaction of interest in the $^{46}$Ar experiment had the largest cross section (elastic scattering), the time spent on Monte Carlo fitting of {\em all} of the events produced in the experiment was the largest computational bottleneck in the analysis. In the case of an experiment where the reaction of interest would represent less than a few percent of the total cross section, this procedure would become highly inefficient and prohibitive. Adding to this the large amount of data produced in this experiment (with even larger data sets expected in future experiments), the analysis simply begs for more efficient analysis tools. 
%Most of the CPU time would be spent on simulating events that are not of interest.
The computationally expensive fitting procedure would be applied to every event, instead of the few percent of the events that are of interest for the analysis.
An unsupervised ML algorithm able to separate the data without {\em a priori} knowledge of the different types of events increases the efficiency of the analysis tremendously, and allows the downstream analysis to concentrate on the fitting efforts only on events of interest. In addition, the clustering allows for more exploration of the data, potentially enabling new discovery of unexpected reaction types.

@@

% The clustering algorithm we found to give the strongest result is the $k$-means algorithm. This algorithm is contingent on having a distance measurement between two datapoints. We found that the standard choice of a euclidian norm on the vector differences provided the best outcomes.

In contrast with the classification work of \citet{Kuchera2019}, we do not assume access to ground truth labels and we are trying to solve a fundamentally different learning problem. Thus, rather than fine-tuning a pre-trained network under the supervised learning regime, we extract the output of the pre-trained network's last convolutional layer as a latent representation of the events, where each event is represented as a vector in $\R^{8192}$. We then cluster events based on this representation using the \texttt{scikit-learn} implementation of the $k$-means algorithm with default parameters \cite{Pedregosa2011}.

%{\color{cyan} [Ryan: I think there are a lot of words in the following two paragraphs that don't need to exist. Essentially, these paragraphs are a repeat of the last paragraph and parts of the last subsection, but with one or two extra details added in (e.g. using the top 100 principle components). As such, I would remove them and add the necessary details to the previous paragraph (or replace them with a small paragraph).]} Building on the work by \citet{Kuchera2019}, we also use the VGG16 pre-trained network as a baseline for the clustering performance. We begin by considering a classical $k$-means approach to clustering. However, the output from the VGG16 network is very high dimensional which traditionally conflicts with good $k$-means results. With output vectors in $\R^{8192}$ the ratio between $L_p$ distances goes to one with increasing dimensionality, as shown by \citet{Aggarwal}. However, one of the central assumptions in the authors' finding is that the elements are uniformly distributed in the latent space. They note that it is then possible that if the class information lies in some sub-space of the latent data, $L_p$-norms may provide useful clusterings even in high dimensional spaces. To investigate this, we perform a $L_2$-norm based $k$-means clustering using the full VGG16 representation and the $10^2$ first principal components only.

%{\color{blue} Are there supposed to be results here? -MPK}
%\subsection{$k$-means}

%We begin by investigating the $k$-means clustering algorithm on the VGG16 latent space. 
% The user a VGG16 model that is pre-trained on the ImageNet data set and subsequently applied to the AT-TPC events without the classification layers. This creates a set of vectors $\boldsymbol{x} \in \R^{8192}$ as our new representation of the events. To cluster, we use the \lstinline{scikit-learn} implementation of the $k$-means++ algorithm with default parameters \cite{Pedregosa2011}.

@@

To compute both the ARI and the clustering accuracy we also have to construct the contingency table between two sets of clustering assignments. A contingency matrix defines the overlaps between classes in these sets, and its general form is shown in Table \ref{tab:contingency}.

@@

\begin{equation}\label{eq:ari}
\text{ARI} = \frac{\sum  \binom{w_{ij}}{2} - \left[\sum  \binom{e_{i}}{2} \sum  \binom{f_{j}}{2}  \right]/\binom{n}{2}}{\frac{1}{2}\left[\sum  \binom{e_{i}}{2} + \sum  \binom{f_{j}}{2}  \right]- \left[\sum  \binom{e_{i}}{2} \sum  \binom{f_{j}}{2}  \right]/\binom{n}{2}}.
\end{equation}
The quantities $f_i$ and $e_i$ are defined in Table \ref{tab:contingency}.
The important distinction to make between the clustering accuracy and the ARI is that the clustering accuracy is a simple comparison that does not account for chance assignments, while the ARI does. In effect, this means that the accuracy is a good heuristic for the performance, tempered by the ARI.

@@

The performance is measured by comparing model predictions on the labelled subset of data (see Table \ref{tab:data sets} in  appendix B for more details).

@@

We first explore the results of applying a $k$-means approach on the latent space of a pre-trained network. Subsequently, we investigate the performance of the MIXAE algorithm as outlined in section \ref{sec:mixae}.

\subsection{$k$-means clustering on the VGG16 latent space}\label{sec:kmeans_results}

The results of the clustering runs are included in Table \ref{tab:clstr_vgg1}. We ran the $k$-means algorithm $N=100$ times with $M=10$ initializations per run, of the cluster centroids to assess the variability in the performance. The $k$-means algorithm returns the best performing model of the $M$ initializations on the unsupervised objective. We report performance on the labelled subset of data, using the labels to identify the top-performing model (Top 1). Additionally, we report the mean and standard deviation of the algorithm on the $N$ trials, which indicate unsupervised performance.

@@

We observe that the clustering on simulated data attains the highest performance, and that there is a decline in performance as we add noise by moving to the filtered and raw data sets. The results are shown in Table \ref{tab:clstr_vgg10}.

\begin{table}[H]
\centering 
\caption[$k$-means on pre-trained model]{$k$-means clustering results on AT-TPC event data in the VGG-16 latent space, for $N=100$ runs of the $k$-means algorithm with $M=10$ initializations. We observe that the performance predictably decreases with the amount of noise in the data.}\label{tab:clstr_vgg10}
\input{./plots/kmeans_vgg_init10.tex}
\end{table}

\noindent The lack of variability is explained by the number of initializations. As can be seen from Table \ref{tab:clstr_vgg1} where we run the $k$-means algorithm $N=1000$ times with $M=1$ initializations of the centroids.

\begin{table}[H]
\centering 
\caption[$k$-means on pre-trained model]{$k$-means clustering results on AT-TPC event data in the VGG-16 latent space, for $N=1000$ runs of the $k$-means algorithm with $M=1$ initializations. We observe that there is significant variability in the results, which is ordinarily masked by $M$ re-initializations that avoid local minima. }\label{tab:clstr_vgg1}
\input{./plots/kmeans_vgg_init1.tex}
\end{table}

In addition to the performance measures reported in Table \ref{tab:clstr_vgg1}, it is interesting to observe which samples are wrongly assigned. To investigate this problem, we compute the matrices $\mathbf{W}$ as shown in Table \ref{tab:contingency}. From these tables, we can infer which classes are more or less entangled with others. The results for each data set is shown in  Figs.~\ref{fig:clster_confmat_sim}, \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. We observe that the proton class is consistently assigned in a pure cluster. For example, consider the row corresponding to the proton class in Fig.~\ref{fig:clster_confmati_filt}. The column corresponding to the largest entry in the proton row has zero other predicted classes in it.

@@

To further investigate the clusters presented in the matrix in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}, we visualize a random subset of examples from the proton events belonging to different clusters for the filtered and full data in Figs.~\ref{fig:filtered_vgg_clster_repr} and \ref{fig:full_vgg_clster_repr}, respectively.
We look at random samples of proton events in two clusters to infer an intuition on the track properties that are considered similar.
The figures indicate that shorter tracks, therefore low-energy or small scattering angle protons, are more likely to appear similar to other event types.
%From these figures, we observe that the more noisy proton events are being clustered together with the amorphous other class events. This is in line with the intuitive expectation from viewing the clustering results in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. A visual inspection of the cluster components serves as a sanity-check for our clustering approach. While we have demonstrated some level of success in clustering events, the algorithm is agnostic to the underlying physics.
    %a major caveat remains in the fundamental aphysical nature of the approach. 
    %The aforementioned figures are then a way to better understand how the algorithm mislabels events of interest. 
    %In lieu of a physics-grounded analysis, we wish to infer the differences which places proton events in other clusters than the high-purity clusters we observe in Figs ~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}.
   % \item While we have achieved some level of success in our clustering methods, we lack the ability to interpret what aspects of the data guided the cluster assignments. To better understand how the algorithm mislabels events of interest, we used visualization of clustered events to infer proton events that the algorithm considers similar. 
%\end{itemize}

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Filteredexamples.pdf}
\caption[Filtered proton samples by cluster belonging]{A sample of proton events from different $k$-means clusters from the filtered data set. The bottom row shows proton samples that are intermingled with noise events in the models' predictions, and the top row belongs to a cluster that contains proton events almost exclusively. Each row belongs to a single cluster corresponding to the filtered confusion matrix in Fig.~\ref{fig:clster_confmati_filt}. While the events in the pure proton cluster are visually distinct from the impure cluster, the difference is less pronounced than what we observe for the full data-set in figure \ref{fig:full_vgg_clster_repr}. }\label{fig:filtered_vgg_clster_repr}.
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Rawexamples.pdf}
\caption[Full proton samples by cluster belonging]{A random sample of proton events from different $k$-means clusters from the raw data set.  Each row belongs to a single cluster corresponding to the full confusion matrix in Fig.~\ref{fig:clster_confmati_raw}. The proton events in the cluster corresponding to the first row appear more visually distinct from other reaction types while the proton events corresponding to the cluster in the bottom row appear visually more similar to other reaction types in our dataset. }\label{fig:full_vgg_clster_repr}
\end{figure}

@@

With the best set of hyperparameters, each highest performing model is thereafter run $10$ times\footnote{The MIXAE model is significantly more computationally expensive to train than a $k$-means model. Resulting in the skew in number of runs in the two cases.}. The results are listed in Table \ref{tab:clstr_mixae}.
We observe that, while the algorithm can achieve a very strong performance, the performance varies. In some cases the MIXAE model converges to a seemingly good configuration, based on its unsupervised training goals. However, when inspecting its clustering performance against labelled data, the seemingly good model does no better than a model based on random selection. 
This happens more frequently with the raw data, indicating an interaction with the noise levels present in the events.

% This table is not properly explained, what is init=10? and what is top 1? 
\begin{table}[H]
\centering 
\caption[MIXAE clustering performance]{MIXAE clustering performance on the ${}^{46} Ar$ experimental data with $N=10$ runs of the algorithm. To quantify the results we report the best performing model (Top 1), and the mean and standard deviation for the result ($\mu\pm \sigma$). In contrast with the VGG-16 + $k$-means approach we observe  significant variations in performance.}\label{tab:clstr_mixae}
\input{./plots/mixae_clustering.tex}
\end{table}
